---
title: "Tidying"
author: "Lauren Zatkos"
date: "May 21, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(dplyr)
library(tidyr)
```
Read in data from the [KNB] (https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.302.1) 
```{r}
catch_df <- catch_df <- read.csv(url("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.302.1", method = "libcurl"),
                    stringsAsFactors = FALSE)
head(catch_df)
```
```{r}
unique(catch_df$notesRegCode)
```
^is a way to check if there is anything in the notes column. And there is, but we won't be analyzing this data (should 
probably go in the metadata for this dataset)

In the below chunk, we selected columns we wanted, specified which columns to name what, and renamed the catch_thousands column. 
```{r}
catch_df <- catch_df %>%
  select(Region, Year, Chinook, Sockeye, Coho, Pink, Chum) %>%
  gather(key = Species, value = Catch, -Region, -Year) %>% 
  rename(catch_thousands = Catch)
head(catch_df)
```  
select specifically picks out columns, while subset picks out rows. 

Question: the table is in wide format, how to we transform it into long format?
?gather
gather turns a dataframe into two columns, you need to specify what you want in the columns/what you don't want (with -)

?spread

If we want to change the catch_thousands value into the actual number of fish. But we got an error because the catch 1000s is currently recognized as character. Because someone accidently entered a letter instead of a number for a chinook count.So we must correct this. 
```{r}
catch_df %>%
  mutate(catch = catch_thousands * 1000)

test<-as.integer(catch_df$catch_thousands)
which(is.na(test) == TRUE)
catch_df[401,]
``` 
So now that we have ID'ed that, this is how to fix it: if else says "if catch thousands = I, make it 1, if not, keep it as whatever value is in catch_thousands"

Found type-o in original data, corrected it, converted catch_thousands column to numeric, and created new column with units # of fish. 
```{r}
catch_df<-catch_df %>%
  mutate(catch_thousands = ifelse(catch_thousands=="I", 1, catch_thousands)) %>%
  mutate(catch_thousands= as.integer(catch_thousands)) %>%
  mutate(catch = catch_thousands * 1000)

```

# Analysis 

```{r}
catch_df %>%
  group_by(Species, Year) %>%
  summarize(mean_catch= mean(catch), num_obs = n())
```

```{r}
catch_df %>%
  filter(Species == "Chinook") %>% 
  group_by(Region) %>% 
  summarise(mean_catch = mean(catch)) %>% #calculate mean catch
  arrange(desc(mean_catch)) #sort in descending order by catch
```